true

### 2020-04-20 analysis/robinson2020.Rmd
debugging latex

---
title: 
    main: Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates
    short: Dendrochronological and radiocarbon dates in the American Southwest
date: "`r format(Sys.Date(), '%B %d, %Y')`"
author:
  - Erick Robinson:
      email: erick.robinson@usu.edu
      institute: [USU]
      correspondence: true
  - Kyle Bocinsky:
      email: kbocinsky@crowcanyon.org
      institute: [CCAC,UM]
      correspondence: false
  - Darcy Bird:
      email: darcy.bird@wsu.edu
      institute: [WSU]
      correspondence: false
  - Jacob Freeman:
      email: jacob.freeman@usu.edu
      institute: [USU]
      correspondence: false
  - Robert L. Kelly:
      email: rlkelly@uwyo.edu 
      institute: [UW]
      correspondence: false
institute:
  - USU: Utah State University
  - CCAC: Crow Canyon Archaeological Center
  - UM: University of Montana
  - WSU: Washington State University
  - UW: University of Wyoming
abstract: >
  The northern American Southwest provides one of the most well-documented cases of human population growth and decline in the world. The geographic extent of this decline in North America is unknown due to the lack of high-resolution palaeodemographic data from regions across and beyond the greater Southwest, where archaeological radiocarbon data is often the only available proxy for investigating these palaeodemographic processes. Radiocarbon time series across and beyond the greater Southwest suggest widespread population collapses from AD 1300--1600. However, radiocarbon data have potential biases caused by variable radiocarbon sample preservation, sample collection, and the non-linearity of the radiocarbon calibration curve. In order to be confident in the wider trends seen in radiocarbon time series across and beyond the greater Southwest, here we focus on regions that have multiple palaeodemographic proxies and compare those proxies to radiocarbon time series. We develop a new method for time series analysis and comparison between dendrochronological data and radiocarbon data. Results confirm a multiple proxy decline in human populations across the Upland US Southwest, Central Mesa Verde, and Northern Rio Grande from AD 1300--1600. These results lend confidence to single proxy radiocarbon-based reconstructions of paleodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.
keywords:
    formatted:
      - dates-as-data
      - radiocarbon summed probability distributions
      - dendrochronology
      - palaeodemography
      - southwestern United States
doi: [DOI HERE]
journal: Preprint
# documentclass: sa
bibliography: robinson2020_references.bib
# csl: "philosophical-transactions-of-the-royal-society-b.csl"
csl: "journal-of-archaeological-science.csl"
link-citations: true
output: 
    bookdown::pdf_document2:
      # template: templates/template.tex
      latex_engine: lualatex
      fig_caption: yes
      toc: false
      extra_dependencies:
        caption: ["labelfont={bf}"]
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
editor_options: 
  chunk_output_type: console
---

@@

```{r Table1}
all_data %>%
  dplyr::bind_rows() %>%
  dplyr::bind_rows(all_data %>%
    dplyr::bind_rows() %>%
    dplyr::mutate(`Study Area` = "UUSW Total")) %>%
  dplyr::filter(!is.na(`Study Area`)) %>%
  dplyr::mutate(`Study Area` = factor(`Study Area`,
    levels = c(
      "UUSW Total",
      "CMV",
      "NRG"
    )
  )) %>%
  dplyr::group_by(Type, SiteID, `Study Area`) %>%
  dplyr::summarise(Dates = dplyr::n()) %>%
  dplyr::group_by(Type, `Study Area`) %>%
  dplyr::summarise(
    Dates = sum(Dates),
    Sites = dplyr::n()
  ) %>%
  tidyr::pivot_wider(names_from = Type, values_from = Dates:Sites) %>%
  dplyr::select(`Study Area`, Dates_Radiocarbon, Sites_Radiocarbon, `Dates_Tree-ring`, `Sites_Tree-ring`) %>%
  knitr::kable(
    caption = "Counts of radiocarbon and tree-ring dates and sites in this study.",
    booktabs = TRUE,
    format.args = list(big.mark = ","),
    col.names = c("Study Area", "Dates", "Sites", "Dates", "Sites")
  ) %>%
  kableExtra::add_header_above(c(" " = 1, "Radiocarbon" = 2, "Tree-ring" = 2))
```

@@

<!-- The following line inserts a page break  -->
\newpage

# References {-}
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

<!-- The following line inserts a page break  -->
\newpage

# Supplementary Material: Cut-width sensitivity for temporal binning {-}
Uneven sampling across archaeological contexts is a persistant problem when constructing summed probability distributions from radiocarbon dates, and perhaps even a larger issue for the more ubiquitous tree-ring dates in regions like the Upland US Southwest. In SPDs of radiocarbon dates, it is conventional to combine dates from single sites prior to creating the SPDs. Dates from separate components or phases of multi-component sites should be counted separately, however.

@@

\newpage

# Colophon {-}

This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017; @rrtools]. This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon-r, cache = FALSE}
# which R packages and versions?
if (requireNamespace("pander", quietly = TRUE)) pander::pander(sessionInfo())
gc()
gc()
```

>>>>

---
title: 
    main: Dendrochronological dates confirm a Late Prehistoric population decline in the American Southwest derived from radiocarbon dates
    short: Dendrochronological and radiocarbon dates in the American Southwest
date: "`r format(Sys.Date(), '%B %d, %Y')`"
author:
  - Erick Robinson:
      email: erick.robinson@usu.edu
      institute: [USU]
      correspondence: true
  - R. Kyle Bocinsky:
      email: kbocinsky@crowcanyon.org
      institute: [CCAC,UM]
      correspondence: false
  - Darcy Bird:
      email: darcy.bird@wsu.edu
      institute: [WSU]
      correspondence: false
  - Jacob Freeman:
      email: jacob.freeman@usu.edu
      institute: [USU]
      correspondence: false
  - Robert L. Kelly:
      email: rlkelly@uwyo.edu 
      institute: [UW]
      correspondence: false
institute:
  - USU: Utah State University
  - CCAC: Crow Canyon Archaeological Center
  - UM: University of Montana
  - WSU: Washington State University
  - UW: University of Wyoming
abstract: >
  The northern American Southwest provides one of the most well-documented cases of human population growth and decline in the world. The geographic extent of this decline in North America is unknown due to the lack of high-resolution palaeodemographic data from regions across and beyond the greater Southwest, where archaeological radiocarbon data is often the only available proxy for investigating these palaeodemographic processes. Radiocarbon time series across and beyond the greater Southwest suggest widespread population collapses from AD 1300--1600. However, radiocarbon data have potential biases caused by variable radiocarbon sample preservation, sample collection, and the non-linearity of the radiocarbon calibration curve. In order to be confident in the wider trends seen in radiocarbon time series across and beyond the greater Southwest, here we focus on regions that have multiple palaeodemographic proxies and compare those proxies to radiocarbon time series. We develop a new method for time series analysis and comparison between dendrochronological data and radiocarbon data. Results confirm a multiple proxy decline in human populations across the Upland US Southwest, Central Mesa Verde, and Northern Rio Grande from AD 1300--1600. These results lend confidence to single proxy radiocarbon-based reconstructions of paleodemography outside the Southwest that suggest post-AD 1300 population declines in many parts of North America.
keywords:
    formatted:
      - dates-as-data
      - radiocarbon summed probability distributions
      - dendrochronology
      - palaeodemography
      - southwestern United States
doi: [DOI HERE]
journal: Preprint
documentclass: sa
bibliography: robinson2020_references.bib
# csl: "philosophical-transactions-of-the-royal-society-b.csl"
# csl: "journal-of-archaeological-science.csl"
csl: "apa.csl"
link-citations: true
output: 
    bookdown::pdf_document2:
      template: templates/template.tex
      latex_engine: lualatex
      fig_caption: yes
      toc: false
      extra_dependencies:
        caption: ["labelfont={bf}"]
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
editor_options: 
  chunk_output_type: console
header-includes:
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}       \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

```{r Table1}
all_data %>%
  dplyr::bind_rows() %>%
  dplyr::bind_rows(all_data %>%
    dplyr::bind_rows() %>%
    dplyr::mutate(`Study Area` = "UUSW Total")) %>%
  dplyr::filter(!is.na(`Study Area`)) %>%
  dplyr::mutate(`Study Area` = factor(`Study Area`,
    levels = c(
      "UUSW Total",
      "CMV",
      "NRG"
    )
  )) %>%
  dplyr::group_by(Type, SiteID, `Study Area`) %>%
  dplyr::summarise(Dates = dplyr::n()) %>%
  dplyr::group_by(Type, `Study Area`) %>%
  dplyr::summarise(
    Dates = sum(Dates),
    Sites = dplyr::n()
  ) %>%
  tidyr::pivot_wider(names_from = Type, values_from = Dates:Sites) %>%
  dplyr::select(`Study Area`, Dates_Radiocarbon, Sites_Radiocarbon, `Dates_Tree-ring`, `Sites_Tree-ring`) %>%
  knitr::kable(
    caption = "Counts of radiocarbon and tree-ring dates and sites in this study.",
    format = "latex",
    booktabs = TRUE,
    format.args = list(big.mark = ","),
    col.names = c("Study Area", "Dates", "Sites", "Dates", "Sites")
  ) %>%
  kableExtra::add_header_above(c(" " = 1, "Radiocarbon" = 2, "Tree-ring" = 2))
```

@@

<!-- The following line inserts a page break  -->
\newpage
# References {-}
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

<!-- The following line inserts a page break  -->
\newpage
\beginsupplement
# Supplementary Material: Cut-width sensitivity for temporal binning {-}
Uneven sampling across archaeological contexts is a persistant problem when constructing summed probability distributions from radiocarbon dates, and perhaps even a larger issue for the more ubiquitous tree-ring dates in regions like the Upland US Southwest. In SPDs of radiocarbon dates, it is conventional to combine dates from single sites prior to creating the SPDs. Dates from separate components or phases of multi-component sites should be counted separately, however.

@@

\newpage
# Colophon {-}
This report is part of a research compendium developed using the `rrtools` package for reproducible research [@Marwick2017; @rrtools]. This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon-r, cache = FALSE}
# which R packages and versions?
sess <- sessionInfo()
sess$basePkgs %<>% sort()
sess$loadedOnly %<>% magrittr::extract(sort(names(sess$loadedOnly)))
if (requireNamespace("pander", quietly = TRUE)) pander::pander(sess)
```
